Hybrid MILP to efficiently and accuratly \\ solve hard DNN verification instances

### 2024-09-24 ICLR25/formula.tex
section 3 and 4

Let $z$ be a neuron of the DNN we want to evaluate, for instance computing an upper bound on the value it can take. We denote by $n< z$ the fact that a neuron appears on a layer before the layer of $z$, and $n \leq z$ if it can also be $z$.
Let us denote ($\Sol\_\max_X^z(n))_{n \leq z}$ is a solution of $\mathcal{M}_X$ 
maximizing $z$. In particular, $\Sol\_\max_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider ($\Sol\_\max_\emptyset^z(n))_{n \leq z}$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
$\Improve\_\max^z(a)=$ $\Sol\_\max_\emptyset^z(z) -$ $\Sol\_\max_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have $\Improve\_\max^z(a)\geq 0$ as $\Sol\_\max_{\{a\}}^z$ fullfils all the constraints of 
$\mathcal{M}_\emptyset$, so $\Sol\_\max_{\{a\}}^z(z) \leq$ $\Sol\_\max_\emptyset^z(z)$.
Similarly, we define ($\Sol\_\min_\emptyset^z(n))_{n \leq z}$ and 
$\Improve\_\min^z(a)$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
Instead, we focus on the following upper bounding of $\Improve\_\max^z(a)$.

@@

For all neurons $n$, let $\sol(n)=$$\Sol\_\max_\emptyset^z(n)$ be the value of neuron $a$
in the solution of the LP instance $\Sol\_\max_\emptyset^z$ to maximize $z$.
%For one layer before the target layer, the formula is simple and most accurate. 
To estimate $\Improve\_\max^z(a)$, we define $\Utility\_\max^z(a)$, first for neurons $a$ one layer before $z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

	\begin{align}
		\Utility\_\max\nolimits^z(a) = |W_{az} \times (\sol(\hat{a})- \ReLU(\sol(a)))|.
	\end{align}

@@

We have that :

\begin{proposition}
$0 \leq \Improve\_\max^z(a) \leq \Utility\_\max^z(a)$. 
\end{proposition}

\begin{proof}
Consider $(\sol'(n))_{n \leq z})_{n \leq z})$ with
$\sol'(n)=\sol(n)$ for all $n \notin \{z,\hat{a}\}$.
In particular,  $\sol'(a) = \sol(a)$.

Now, define $\sol'(\hat{a}) = \ReLU(\sol(a))$. 
That is,  
$\sol'(\hat{a})$ is the correct value for $\hat{a}$ (obtained if we open neuron $a$), 
compared to the LP abstraction for $\sol(\hat{a})$.

Last, we define $\sol'(z)=\sol(z) + W_{az} \times (\ReLU(\sol(a)) - \sol(\hat{a})) = \sol(z) - \Utility\_\max^z(a)$.

Now, it is easy to check that $(\sol'(n))_{n \leq z}$ satisfies the constraints in 
$\mathcal{M}_{\{a\}}$, as opening $a$ changes the value of $\hat{a}$ from
$\sol(\hat{a})$ to $\ReLU(\sol(a))$, and the contribution from $a$ to $z$ is 
$W_{az}$. As $\sol'(z)$ is a solution of $\mathcal{M}_{\{a\}}$, it is smaller or equal to the maximal solution: $\sol'(z) \leq$ $\Sol\_\max_{\{a\}}^z(z)$. That is, 
$\sol(z)-\sol'(z) \geq \sol(z) -$ $\Sol\_\max_{\{a\}}^z(z)$, i.e. 
$\Improve\_\max^z(a) \leq \Utility\_\max^z(a)$.
\end{proof}

Thus, $\Utility\_\max^z(a)$ can be used to approximate 
$\Improve\_\max^z(a)$. In particular, consider a very frequent case where $W_{az}<0$.
In this case, to maximize $z$, the LP engine sets $\sol(\hat{a})$ to its minimal value\dots which happens to be $\sol(\hat{a})=\ReLU(a)$, thanks to Proposition \ref{LP}. 
In this frequent case, we have $\Utility\_\max^z(a)=0$, and thus $\Improve\_\max^z(a)=0$. 
In particular, this node will have the smallest utility (thus will not get picked in the open nodes $X$), and indeed it is not having any impact on  $\Sol\_\max_{\{a\}}^z(z)$. 
This is one stricking difference (but not the only one) with choosing utility based on 
$|W_{az}|$ \cite{DivideAndSlide}.

@@

Now, consider neurons $a$ two layers before $z$. We use $b$ to denote neurons in the layer between $a$ and $z$. If we open $a$, then the change in the weight of $a$ will still be $\sol(\hat{a})-\ReLU(\sol(a))$ as above. Its impact on $z$ is no more direct with $W_{az}$, but it is through all the $b$ in the intermediate layer. 
Based on Proposition \ref{LP}, we can evaluate how opening $a$ impacts each value of 
$\hat{b}$,  by using the upper and lower bound
$\UB(b),\LB(b)$.

Indeed, if $W_{bz}\leq0$, then the LP solver sets $\sol(\hat{b})$ to the lowest possible value to maximize $z$, which happens to be $\ReLU(b)$ according to Proposition \ref{LP}.
If $\sol(b) < 0$, then we have a clip at $\sol(\hat{b})=\ReLU(b)=0$ and opening $a$ does not help. If $\sol(b) > 0$, then we have $\sol(\hat{b})=\ReLU(b)=\sol(b)$ and every change from 
$a$ is reflected with a coefficient $W_{ab}$ on $\hat{b}$. 
Last, if $W_{bz}>0$, then according to Proposition \ref{LP}, the LP solver
sets $\sol(\hat{b}) = \sol(b) \frac{\UB(b)}{\UB(b)-\LB(b)}$ to maximize $z$.

@@

We define a difference $\Delta$ to account for the fact that a neuron $b$ 
can not exceed $\LB(b),\UB(b)$:
	\begin{align*}
		&\Delta(\hat{a}) = \sol(\hat{a}) - \ReLU(\sol(a)) \geq 0\\
		&\Delta(b) = \begin{cases}
			\max(W_{ab}\Delta(\hat{a}),\LB(b)-\sol(b)),  &\text{ if }  W_{ab} < 0 \\
			\min(W_{ab}\Delta(\hat{a}),\UB(b)-\sol(b)),  &\text{ if }  W_{ab} > 0 \\
		\end{cases}\\
			&\Delta(\hat{b}) =
		\begin{cases}
			\max(\Delta(b),-\sol(b)),  &\text{if }  W_{bz} > 0 \text{ and } \sol(b)\geq0\\
				\max(\Delta(b)+\sol(b),0),  &\text{if }  W_{bz} > 0 \text{ and } \sol(b)<0\\
		 \frac{\UB(b)}{\UB(b)-\LB(b)}\Delta(b),  &\text{if }W_{bz} < 0\\
		\end{cases}
		\end{align*}

We can now set 
$$ \Utility\_\max\nolimits^z(a) = \sum_b W_{bz} \Delta(\hat{b}).$$

We proceed inductively in the same way to define $\Utility\_\max^z(a)$ for deeper neuron $a$.

>>>>

Let $z$ be a neuron of the DNN we want to evaluate, for instance computing an upper bound on the value it can take. We denote by $n < z$ the fact that a neuron $n$appears on a layer before the layer of $z$, and $n \leq z$ if it can also be $z$.
Let us denote ($\Sol\_\max_X^z(n))_{n \leq z}$ a solution of $\mathcal{M}_X$ 
maximizing $z$. In particular, $\Sol\_\max_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider ($\Sol\_\max_\emptyset^z(n))_{n \leq z}$ (this can be obtained very efficiently by {\em one} call to an LP solver). An accurate utility function is to evaluate 
$\Improve\_\max^z(a)=$ $\Sol\_\max_\emptyset^z(z) -$ $\Sol\_\max_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have $\Improve\_\max^z(a)\geq 0$ as $\Sol\_\max_{\{a\}}^z$ fullfils all the constraints of 
$\mathcal{M}_\emptyset$, so $\Sol\_\max_{\{a\}}^z(z) \leq$ $\Sol\_\max_\emptyset^z(z)$.
Similarly, we define ($\Sol\_\min_\emptyset^z(n))_{n \leq z}$ and 
$\Improve\_\min^z(a)$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
Instead, we focus on the following upper bounding of $\Improve\_\max^z(a)$.

@@

For all neurons $n$, let $\sol(n)=$$\Sol\_\max_\emptyset^z(n)$ be the value of neuron $a$
in the solution of the LP instance $\Sol\_\max_\emptyset^z$ to maximize $z$.
%For one layer before the target layer, the formula is simple and most accurate. 
To estimate $\Improve\_\max^z(a)$, we define $\Utility\_\max^z(a)$, first for neurons $a$ one layer before $z$, by computing by how much the value of $z$ will change if $a$ is opened
and other values remain the same - in particular, $value(\hat{a})=\ReLU(sol(a))$. We define:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

	\begin{align}
		\Utility\_\max\nolimits^z(a) = W_{az} \times (\sol(\hat{a})- \ReLU(\sol(a))).
	\end{align}

@@

In the case $W_{az}<0$, to maximize $z$, the LP engine sets $\sol(\hat{a})$ to its minimal value, which is $\sol(\hat{a})=\ReLU(a)$ thanks to Proposition \ref{LP}. In this case, we have $\Utility\_\max^z(a)=0$.
Thus, we have $\Utility\_\max^z(a) \geq 0$ in all cases
as $\sol(\hat{a}) \geq \ReLU(a)$. 
We will show with a more general definition that $0 \leq \Improve\_\max^z(a) \leq \Utility\_\max^z(a)$ in Prop.~\ref{prop2}. Thus, $\Utility\_\max^z(a)$ can be used to approximate $\Improve\_\max^z(a)$. In particular, for all nodes $a$ with $W_{az}< 0$,
this node will have the smallest $\Utility\_\max\nolimits^z(a)=0$ (thus will not get picked in the open nodes $X$), and indeed it is not having any impact on  $\Sol\_\max_{\{a\}}^z(z)$. This is one stricking difference (but not the only one) with choosing utility based on 
$|W_{az}|$ \cite{DivideAndSlide}.

@@

Now, consider neurons $a$ two layers before $z$. We use $b$ to denote neurons in the layer $\ell$ between $a$ and $z$. 
If we open $a$, then the change $\Delta(a)$ in the weight of $a$ is 
$\Delta(a)=\ReLU(\sol(a)) - \sol(\hat{a}) \leq 0$ as above. Its impact on $z$ is no more direct with $W_{az}$, but it is through all the $b \in \ell$. 
We let $\Delta(b) = W_{ab}\Delta(\hat{a})$.
Based on Proposition \ref{LP}, we can evaluate the impact 
$\Delta(\hat{b})$ of opening $a$ on each value of $\hat{b}$, by using the upper and lower bound $\UB(b),\LB(b)$:

	\begin{align*}
		&\Delta(\hat{b}) =
		\begin{cases}
			\frac{\UB(b)}{\UB(b)-\LB(b)}\Delta(b),  &\text{if }W_{bz} > 0\\
			\max(\Delta(b),-\sol(b)),  &\text{if }  W_{bz} < 0 \text{ and } \sol(b)\geq0\\
			\max(\Delta(b)+\sol(b),0),  &\text{if }  W_{bz} < 0 \text{ and } \sol(b)<0		 
		\end{cases}
		\end{align*}

Indeed, if $W_{bz}>0$, then according to Proposition \ref{LP}, the LP solver
sets $\sol(\hat{b}) = \sol(b) \frac{\UB(b)}{\UB(b)-\LB(b)} +$ Cst to maximize $z$.
Changing $b$ by $\Delta(b)$ thus results in changing $\sol(\hat{b})$ by 
$\frac{\UB(b)}{\UB(b)-\LB(b)}\Delta(b)$.
If $W_{bz}\leq0$, then the LP solver sets $\sol(\hat{b})$ to the lowest possible value to maximize $z$, which happens to be $\ReLU(b)$ according to Proposition \ref{LP}.
If $\sol(b) < 0$, then we have $\sol(\hat{b})=\ReLU(b)=0$ and opening $a$ change the 0 value only if $\sol(b)+\Delta(b)>0$. If $\sol(b) > 0$, then 
$\sol(\hat{b})=\ReLU(\sol(b))=\sol(b)$, and the change to $\hat{b}$ will be 
the full $\Delta(b)$, unless $\Delta(b) < -\sol(b) < 0$ in which case it is 
$-\sol(b)$. We then set:

@@

$$ \Utility\_\max\nolimits^z(a) = -\sum_{b \in \ell} W_{bz} \Delta(\hat{b})$$

\begin{proposition}
	\label{prop2}
		$0 \leq \Improve\_\max^z(a) \leq \Utility\_\max^z(a)$. 
	\end{proposition}

	\begin{proof}
    Consider $\sol'(n))_{n \leq z})$ with
	$\sol'(n)=\sol(n)$ for all $n \notin \{z,\hat{a}\} \cup \{b,\hat{b} \mid b \in \ell\}$. In particular,  $\sol'(a) = \sol(a)$.
	Now, define $\sol'(\hat{a}) = \ReLU(\sol(a))$. 
	That is, $\sol'(\hat{a})$ is the correct value for $\hat{a}$, obtained if we open neuron $a$, compared to the LP abstraction for $\sol(\hat{a})$.
	We also define $\sol'(b)=\sol(b)+\Delta(b)$ and 
	$\sol'(\hat{b})=\sol(\hat{b}) + \Delta(\hat{b})$.
	Last, $\sol'(z)=\sol(z) + \sum_{b \in \ell} W_{bz} \Delta(\hat{b})$.
	% = \sol(z) - \Utility\_\max\nolimits^z(a)$.

	It is easy to check that $(\sol'(n))_{n \leq z}$ satisfies the constraints in 
	$\mathcal{M}_{\{a\}}$, as opening $a$ changes the value of $\hat{a}$ from
	$\sol(\hat{a})$ to $\ReLU(\sol(a))$, and the contribution from $a$ to $b,\hat{b},z$ 
	are respected.
	As $\sol'(z)$ is a solution of $\mathcal{M}_{\{a\}}$, it is smaller or equal to the maximal solution: $\sol'(z) \leq$ $\Sol\_\max_{\{a\}}^z(z)$. That is, 
	$\sol(z)-\sol'(z) \geq \sol(z) -$ $\Sol\_\max_{\{a\}}^z(z)$, i.e. 
	$\Improve\_\max^z(a) \leq \Utility\_\max^z(a)$.
	In particular, we have that $\Utility\_\max^z(a) \geq 0$, which was not obvious from the definition.
	\end{proof}

We can proceed inductively in the same way to define $\Utility\_\max^z(a)$ for deeper neurons $a$.

### 2024-09-24 ICLR25/valabstraction.tex
update
section 3 and 4

%\iffalse
%An extremely efficient algorithm that alleviates some of the inaccuracies is DeepPoly \cite{deeppoly}, also independently discovered as the CROWN algorithm \cite{crown}. Instead of fixed bounds for each neuron $n$ in layer $k$, DeepPoly maintains two affine functions representing the lower and upper bounds of the neuron's value, based on inputs from the previous layer $k-1$. For instance, denote $f_i \leq x_i \leq g_i$ with, for example, $f_{3}(x_3)=f_4(x_4)=0$ and $g_3(x_3) = \frac{x_3+2}{2}$, $g_4(x_4) = \frac{x_4+2}{2}$. This leads to $x_5 \leq g_3(x_3) + 2 g_4(x_4) = \frac{x_3 + 2x_4 + 6}{2} = \frac{3x_1 - x_2 + 6}{2}\leq 5$. For bounds $[\UB,\LB]$ on $x_i$, the optimal linear function for the upper bound is $g_i(x_i)=\LB \frac{x_i-\alpha}{\LB-\UB}$ for ReLU nodes. There are two options for the lower bound: $f^1_i(x_i) = 0$ or $f^2_i(x_i)=x_i$. DeepPoly selects between these based on the values of $\alpha$ and $\beta$ (for unstable neurons): if $|\alpha|\geq |\beta|$, then $f_i=f^1_i$ is chosen, otherwise, for $|\beta|>|\alpha|$, $f_i=f^2_i$ is selected. 
%\fi

@@

At the other end of the spectrum, we find the Mixed Integer Linear Programming (MILP) value abstraction, which is a complete (but inefficient) method. 
Consider an unstable neuron $n \in[\UB,\LB]$. The value $\hat{x}$ of $\ReLU(x)$ can be encoded exactly in an MILP formula with one integer (actually even binary) variable $a$ valued in $\{0,1\}$, using constants $(\UB,\LB)$ with 4 constraints \cite{MILP}:
$\quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \LB \cdot a \quad \wedge \quad \hat{x} \leq x-\UB \cdot (1-a)$.

For all $x \in [\UB,\LB] \setminus 0$, there exists a unique solution $(a,\hat{x})$ that meets these constraints, with $\hat{x}=\ReLU(x)$ \cite{MILP}. Here, $a$ is 0 if $x < 0$, 1 if $x>0$, and can be either if $x=0$~\cite{MILP}. This encoding approach can be applied to every (unstable) ReLU node, and optimizing its value can help getting more accurate bounds. However, for networks with hundreds of {\em unstable} nodes, the resulting MILP formulation will contain numerous integer variables and generally bounds obtained will not be accurate, even using powerful commercial solvers such as Gurobi.

@@

Consider an unstable neuron $n$, that is $\LB(n) < 0 < \UB(n)$. 
Let $x$ represent the input value of $n$, hence with 
$\LB(n) < x < \UB(n)$, and $\hat{x}$ representing $ReLU(x)$. 
Then we have:

\begin{proposition}
	\label{LP}
	The LP relaxation on $\hat{x}$ is equivalent with:
	\begin{align}
		\ReLU(x) \leq \hat{x} \leq \UB(n) \frac{x-\LB(n)}{\UB(n)-\LB(n)} \label{eq:deeppoly}
		%& \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \beta \cdot a \quad \wedge \quad \hat{x} \leq x-\alpha \cdot (1-a), \, \quad a \in [0,1]  \label{eq:lp}\\
	\end{align} 
\end{proposition}

>>>>

@@

At the other end of the spectrum, we find the Mixed Integer Linear Programming (MILP) value abstraction, which is a complete (but inefficient) method. 
Consider an unstable neuron $n$, whose value $x \in [\LB(n),\UB(n)]$ with $\LB(n)<0<\UB(n)$.
The value $\hat{x}$ of $\ReLU(x)$ can be encoded exactly in an MILP formula with one 
integer (actually even binary) variable $a$ valued in $\{0,1\}$, using constants $\UB(n),\LB(n)$ with 4 constraints \cite{MILP}:

$$\quad \hat{x} \geq x \quad \wedge \quad \hat{x} \geq 0, \quad \wedge \quad \hat{x} \leq \LB(n) \cdot a \quad \wedge \quad \hat{x} \leq x-\UB(n) \cdot (1-a)$$

For all $x \in [\LB(n),\UB(n)] \setminus 0$, there exists a unique solution $(a,\hat{x})$ that meets these constraints, with $\hat{x}=\ReLU(x)$ \cite{MILP}. The value of $a$ is 0 if $x < 0$, and 1 if $x>0$, and can be either if $x=0$. This encoding approach can be applied to every (unstable) ReLU node, and optimizing its value can help getting more accurate bounds. However, for networks with hundreds of {\em unstable} nodes, the resulting MILP formulation will contain numerous integer variables and generally bounds obtained will not be accurate, even using powerful commercial solvers such as Gurobi.

@@

Consider an unstable neuron $n$, that is $\LB(n) < 0 < \UB(n)$. 
Let $x$ represent the input value of $n$, hence with 
$\LB(n) < x < \UB(n)$, and $\hat{x}$ representing $\ReLU(x)$. 
We have:

\begin{proposition}
	\label{LP}
	The LP relaxation on $\hat{x}$ is equivalent with:
	\begin{align}
		\ReLU(x) \leq \hat{x} \leq \UB(n) \frac{x-\LB(n)}{\UB(n)-\LB(n)} \label{eq:deeppoly}
	\end{align} 
\end{proposition}
